The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-15 paper.tex
...

\author{
Francisco Sant'Anna
\affil{PUC--Rio}
Adriano Branco
\affil{PUC--Rio}
Roberto Ierusalimschy
\affil{PUC--Rio}
Noemi Rodriguez
\affil{PUC--Rio}
Silvana Rossetto
\affil{UFRJ}
}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Its synchronous programming model allows for a simple reasoning about 
shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.

In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
code dissemination in sensor networks.
\end{abstract}

@@

- Esterel

- Ceu
    - first WSN
    - other domains, such as games
    - here, only static semantics
    - uses, classes, gsoc,

- synchronous limitations

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel and 
\CEU for the following control specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification is translated almost identically in the 
two languages, as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent (lines 4--9).
%
For the second phrase (i.e., the reset behavior), the Esterel version uses a 
\code{abort-when}, which serves the same purpose of \CEU's \code{par/or}:
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The same specification in Esterel and \CEU. %\newline
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\CEU (like Esterel) has a strong imperative flavor, with explicit control flow 
through sequences, loops, and also assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
Programs advance in sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

In the sections that follow, we discuss the four basic differences between \CEU 
and Esterel: deterministic execution for operations with side-effects
(Section~\ref{sec.ceu.det}), safe abortion for lines of execution 
(Section~\ref{sec.ceu.abrt}), first-class timers 
(Section~\ref{sec.ceu.timers}), and stack-based execution for internal events 
(Section~\ref{sec.ceu.ints}).

@@

Esterel is deterministic with respect to reactive control: ``the same sequence 
of inputs always produces the same sequence of outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in ``\code{call 
f1() || call f2()}'', the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, as in
``\code{par/and~do~\_f1()~with~\_f2()~end}'', they are scheduled in the order 
they appear in the program source code (i.e., \code{\_f1} executes first).
%
This way, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency possible.
In contrast, Esterel does not support shared memory: \emph{``if a variable is 
written by some thread, then it can neither be read nor be written by 
concurrent threads''}~\cite{esterel.primer}.
%
For embedded development, we believe that deterministic shared-memory 
concurrency is beneficial, given the extensive use of memory mapped ports for 
I/O and lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The execution order between \code{f1} and \code{f2} is unspecified in 
Esterel, whereas in \CEU, \code{\_f1} always executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\subsection{Safe Abortion}
\label{sec.ceu.abrt}

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution (i.e., awaiting \code{A} and 
\code{B}) without tweaking them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.

The code fragments of Figure~\ref{lst.abortion} shows a corner case for 
abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
%
For instance, it is not clear in \emph{code~a} in Esterel if the call to 
\code{f} should execute or not after \code{A}, given that the body and abortion 
events are the same.
%
For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
the \code{abort} statement.
With \emph{strong} abortion (the default), the body is aborted immediately and 
the call does not execute.
%
In \CEU, given the deterministic scheduling rules, strong and weak abortions 
can be chosen by reordering trails inside a \code{par/or}, e.g., in 
\emph{code~b}, the second trail is strongly aborted by the first trail and the 
call to \code{\_f} never executes.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
// CEU
par/or do
    await A;
with
    await A;
    _f();
end
\end{lstlisting}
\end{minipage}
%
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Thread abortion in Esterel and \CEU. %\newline
\label{lst.abortion}
}
\end{figure}

@@

% TODO: ana

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of WSNs, such as timeouts and sensor sampling.
However, support for wall-clock time is somewhat low-level in existing 
languages, usually through timer callbacks or blocking calls with ``sleep''.
%
In any concrete system implementation, however, a requested timeout does not 
expire precisely with zero-delay, a fact that is usually ignored in the 
development process.
We define the difference between the requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in 
sequence (or in a loop) may accumulate a considerable amount of deltas that can 
lead to incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
As an example, consider the following program:

\nopagebreak
\noindent
\begin{minipage}{\linewidth}
\begin{lstlisting}[xleftmargin=2em]
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}

Suppose that after the first \code{await} request, the underlying system gets 
busy and takes 15ms to check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program that follows, although the scheduler cannot 
guarantee that the first trail terminates exactly in 11ms, it can at least 
ensure that the program always terminates with \code{v=1}:

\nopagebreak
\noindent
%\begin{figure}[t]
%\rule{8.5cm}{0.37pt}
\begin{minipage}{\linewidth}
\begin{lstlisting}[xleftmargin=2em]
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
%\caption{ The first trail always terminates the program.
%\label{lst.time}
%}
%\end{figure}

Remember that any non-awaiting sequence is considered to take no time in the 
synchronous model.
Hence, the first trail is guaranteed to terminate before the second trail, 
because $10+1 < 12$.
A similar program in a language without first-class support for timers, would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.
%
The importance of synchronized timers becomes more evident in the presence of 
loops, like in the introductory example of Figure~\ref{lst.all} in which the 
first trail is guaranteed to execute exactly 21 times before being aborted by 
the timer in the second trail.

Note that in extreme scenarios, small timers in sequence (or in a loop) may 
never ``catch up'' with the external clock, resulting in a \emph{delta} that
increases indefinitely.
%There is nothing \CEU can do about that, as it is do
To deal with such cases, the current \emph{delta} is always returned from an 
\code{await} and can be used in programs:

\begin{lstlisting}
loop do
    var int late = await 1ms;
    if late < 1000 then
        <...>   // normal behavior
    else
        <...>   // abnormal behavior
    end
end
\end{lstlisting}

% TODO: uses in ports
% trickle?

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
For the version in Esterel, given that there is no control dependency between 
the calls to \code{f}, they may execute in any order after \code{A} and 
\code{B} (internally emitted).
%
For the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st]}
\item 2nd trail awakes (line 9), calls \code{\_f(1)}, and terminates.\\
    \emph{stack: [1st]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f(2)}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}[!t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f("2");
$\|$
    await B;
    call f("1");
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external
event void b;  // internal
par/and do
    await A;
    emit b;
    _f("2");
with
    await b;
    _f("1");
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself will always be 
ignored, given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring runtime stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, just like coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.
%
In Section~\ref{sec.adv.frp}, we take advantage of the lack of recursion to 
properly describe mutual dependency among trails in parallel.

@@

The disciplined synchronous execution of \CEU, together with broadcast 
communication and stacked execution for internal events, may raise doubts about 
the precise execution of programs.
%
In this chapter, we introduce a reduced syntax of \CEU and propose an 
operational semantics in order to formally describe the language, eliminating 
imprecisions with regard to how a program reacts to an external event.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

@@

\begin{figure}[h]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\end{minipage}
%
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\end{minipage}
%
\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        u8 _and_3: 1;
        u8 _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

>>>>

\author{
Francisco Sant'Anna
\affil{Departamento de Inform\'atica, PUC--Rio}
Adriano Branco
\affil{Departamento de Inform\'atica, PUC--Rio}
Roberto Ierusalimschy
\affil{Departamento de Inform\'atica, PUC--Rio}
Noemi Rodriguez
\affil{Departamento de Inform\'atica, PUC--Rio}
Silvana Rossetto
\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms.
%
Employing the synchronous programming model, it allows for a simple reasoning 
about shared-memory concurrency.
Furthermore, its restricted semantics enables deterministic and memory-safe 
programs.
%
% TODO: some uses?
%
In this work, we propose a formal semantics for \CEU focusing on its particular 
control mechanisms, such as parallel compositions, finalization, and 
stack-based internal events.
%
We also present an implementation with two backends:
one aiming for resource efficiency and interoperability with C, and another for 
code dissemination in sensor networks.
\end{abstract}

@@

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Movendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
\end{itemize}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel and 
\CEU for the following control specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

\begin{figure}[t]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
%\newline
{\small
%TODO.
}
\label{lst.abro}
}
\end{figure}

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In the synchronous model, programs advance in a sequence of discrete reactions 
to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
native calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.

In the sections that follow, we review the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}: deterministic execution 
(Section~\ref{sec.ceu.det}), safe abortion with finalization 
(Section~\ref{sec.ceu.abrt}), first-class timers 
(Section~\ref{sec.ceu.timers}), and stack-based internal events 
(Section~\ref{sec.ceu.ints}).

@@

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{"call 
f1() || call f2()"}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
In \CEU, when multiple trails are active at a time, they are scheduled in the 
order they appear in the program source code.
%
Hence, \CEU is deterministic also with respect to the order of execution of 
side effects within a reaction.
%
Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to illustrate the semantic difference regarding 
(non-)determinism.

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
In contrast, Esterel precludes support for shared memory: \emph{``if a variable 
is written by some thread, then it can neither be read nor be written by 
concurrent threads''}~\cite{esterel.primer}.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other embedded languages made a similar design choice~\cite{wsn.sol,pret}.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.det}
}
\end{figure}

\subsection{Safe Abortion with Finalization}
\label{sec.ceu.abrt}

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
However, handling abortion when dealing with external resources can be 
challenging, given that external entities are not subject to the same 
synchronous execution discipline.

\begin{figure}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{A network protocol in \CEU extended with a finalization clause to 
cancel a transmission.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

The example in the left of Figure~\ref{lst.fin} relies on hierarchical 
\code{par/or} and \code{par/and} compositions to describe the state machine of 
a data collection protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has two trails:
one monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles an immediate retransmission request (line 7);
the other transmits the packet awaiting for a confirmation (lines 9--16).
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Note that the \code{buffer} packet (line 12) is a local variable whose address 
is passed to function \code{\_send\_enqueue} for transmission.
The call enqueues the pointer in the radio driver, which holds it up to the 
emission of \code{SENDACK} acknowledging the packet transmission.
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

%Programs in \CEU can access \emph{C} libraries available in the underlying 
%platform by prefixing symbols with an underscore (e.g., 
%\code{\_printf(<...>)}).
%
The \CEU compiler tracks the interaction of \code{par/or} compositions with 
local variables and stateful \emph{C} functions (e.g., device drivers) in order 
to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
For instance, \CEU refuses to compile the program in the left of 
Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
clause to accompany the stateful \emph{C} call~\cite{ceu.sensys13}.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

\begin{comment}
The code fragments of Figure~\ref{lst.abortion} shows a corner case regarding
abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
%
For instance, it is not clear in the code in Esterel if the call to \code{f} 
should execute or not after \code{A}, given that the body and abortion events 
are the same.
%
For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
the \code{abort} statement.
With \emph{strong} abortion (the default), the body is aborted immediately and 
the call does not execute.
%
In \CEU, given the deterministic scheduling rules, strong and weak abortions 
can be chosen by reordering trails inside a \code{par/or}, e.g., in 
Figure~\ref{lst.abortion}, the second trail is strongly aborted by the first 
trail and the call to \code{\_f} never executes.

\begin{figure}[t]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// ESTEREL
abort
    await A;
    call f();
when A;

\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
// CEU
par/or do
    await A;
with
    await A;
    _f();
end
\end{lstlisting}
\end{minipage}
%
%\rule{8.4cm}{0.37pt}
\caption{ Strong abortion in Esterel and \CEU. %\newline
\label{lst.abortion}
}
\end{figure}
\end{comment}

@@

Activities that involve reactions to \emph{wall-clock time}%
\footnote{
By wall-clock time we mean the passage of time from the real world, measured in 
hours, minutes, etc.
}
appear in typical patterns of embedded development, such as timeouts and sensor 
sampling.
However, support for wall-clock time is somewhat low-level in existing 
languages, usually through timer callbacks or ``sleep'' blocking calls.
%
However, in any concrete system implementation, a requested timeout does not 
expire precisely with zero-delay, a fact that is usually ignored in the 
development process.
We define the difference between the requested timeout and the actual expiring 
time as the \emph{residual delta time (delta)}.
Without explicit manipulation, the recurrent use of timed activities in 
sequence (or in a loop) may accumulate a considerable amount of deltas that can 
lead to incorrect behavior in programs.

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
var int v;
await 10ms;
v = 1;
await 1ms;
v = 2;
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}
par/or do
    await 10ms;
    <...>         // any non-awaiting sequence
    await  1ms;
    v = 1;
with
    await 12ms;
    v = 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{ First-class timers in \CEU.
\label{lst.timers}
}
\end{figure}

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}:
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers, would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they are both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st]}
\item 2nd trail awakes (line 9), calls \code{\_f(1)}, and terminates.\\
    \emph{stack: [1st]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f(2)}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, just like coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

%The disciplined synchronous execution of \CEU, together with broadcast 
%communication and stacked execution for internal events, may raise doubts 
%%about the precise execution of programs.
%
In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics in order to formally describe the language.
%, eliminating imprecisions with regard to how a program reacts to an external 
%event.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and $C$ calls (which behave like in any conventional 
imperative language).

@@

\begin{figure}[h]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\end{minipage}
%
\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\begin{figure}[t]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        u8 _and_3: 1;
        u8 _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}
